---
title: "ccImpute package manual"
author: "Marcin Malec"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{ccImpute package manual}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: references.bib  

---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```
# Introduction
Dropout events make the lowly expressed genes indistinguishable from true zero 
expression and different than the low expression present in cells of the same
type. This issue makes any subsequent downstream analysis difficult. 
ccImpute[@malec2022ccimpute] isan imputation tool that uses cell similarity 
established by consensus clustering to impute the most probable dropout events
in the scRNA-seq datasets. ccImpute demonstrates performance which exceeds the 
performance of existing imputation approaches while introducing the least amount
of new noise as measured by clustering performance characteristics on datasets 
with known cell identities.

## Installation.
```{r  eval=FALSE}
To install this package, start R (version "4.2") and enter:
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ccImpute")
```

# Data Pre-processing
`ccImpute` is an imputation tool and it does not provide functions for the
pre-processing the data. This tool expects the user to preprocess the data
prior to using it. The input data is expected to be in log-normalized format.
This manual includes sample minimal pre-processing of dataset from
[scRNAseq database](http://bioconductor.org/packages/scRNAseq) using the
[scater tool](http://bioconductor.org/packages/scater).

# Sample Usage
## Required libraries
```{r setup, message=FALSE, warning=FALSE}
library(scRNAseq)
library(scater)
library(ccImpute)
library(SingleCellExperiment)
library(stats)
library(mclust)
```

## Input Data.
The following code loads Darmanis dataset[@darmanis2015survey] and computes
log-transformed normalized counts:
```{r message=FALSE, warning=FALSE}
sce <- DarmanisBrainData()
sce <- logNormCounts(sce)
```
## Pre-processing data.
A user may consider performing [feature selection]
(https://bioconductor.org/books/3.15/OSCA.basic/feature-selection.html)
prior to running the imputation. ccImpute only imputes the most probable 
dropout events and is unlikely to benefit from presence of scarcely expressed
genes nor make any corrections to their expression.


## Adjusted Rand Index (ARI)
Adjusted Rand Index is a measure of the similarity between two data clusterings
that is adjusted for the chance grouping of elements. This measure allows to
evaluate performance of clustering algorithm as a similarity to the optimal 
clustering assignments derrived from cell labels.

 
## Compute Adjusted Rand Index (ARI) without imputation.
```{r}
# Set seed for reproducibility purposes.
set.seed(0) 
# Compute PCA reduction of the dataset
reducedDims(sce) <- list(PCA=prcomp(t(logcounts(sce)))$x)

# Get an actual number of cell types
k <- length(unique(colData(sce)$cell.type))

# Cluster the PCA reduced dataset and store the assignments
set.seed(0) 
assgmts <- kmeans(reducedDim(sce, "PCA"), centers = k, iter.max = 1e+09,
    nstart = 1000)$cluster

# Use ARI to compare the k-means assignments to label assignments
adjustedRandIndex(assgmts, colData(sce)$cell.type)
```
## Perform the imputation and update the logcounts assay.
```{r}
assay(sce, "imputed") <- ccImpute(assays(sce)$logcounts, k = k)
```
## Recompute Adjusted Rand Index (ARI) with imputation.
```{r}
# Recompute PCA reduction of the dataset
reducedDim(sce, "PCA_imputed") <- prcomp(t(assay(sce, "imputed")))$x

# Cluster the PCA reduced dataset and store the assignments
assgmts <- kmeans(reducedDim(sce, "PCA_imputed"), centers = k, iter.max = 1e+09,
    nstart = 1000)$cluster

# Use ARI to compare the k-means assignments to label assignments
adjustedRandIndex(assgmts, colData(sce)$cell.type)
```

# R session information.
```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```
